---
title: "lecture 13"
author: "Julia Olivieri"
date: "2023-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data set

In this report we'll be analyzing data from the 2015 World Happiness Report.

We'll start by reading in the data:

```{r}
library(tidyverse)

countries <- read_csv("~/Desktop/COMP_162/datasets/happiness_2015.csv")
```
## Exploratory data analysis

### Descriptive statistics

Let's start by finding some summaries of our variables.

```{r}
summary(countries)
```
Based on this output, we can see that our quantitative variables (other than `HappinessScore`) seem to fall between 0 and 2 (so for example `LifeExpectancy`  isn't in years).

We have two categorical variables. Let's see how many observations of each type we have for these.

```{r}
countries %>%
  group_by(Country) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```
Looks like there's only one entry for each country, just as we'd suspect. Let's do the same for region:

```{r}
countries %>%
  group_by(Region) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```

We can see that the five regions with the greatest number of countries contain the vast majority of countries. This could be useful if we want to subset to only specific regions later in our analysis.

Let's break down some of these scores more. I'm interested in finding how the means of our various quantitative variables vary by region.

```{r}
countries %>%
  group_by(Region) %>%
  summarize(avg_score = mean(Generosity)) %>%
  arrange(desc(avg_score))
```

```{r}
countries %>%
  group_by(Region) %>%
  summarize(avg_score = mean(GDPperCapita)) %>%
  arrange(desc(avg_score))
```

```{r}
countries %>%
  group_by(Region) %>%
  summarize(avg_score = mean(Family)) %>%
  arrange(desc(avg_score))
```

### Exploratory visualizations

Now let's look at some visualizations of our data. Let's look at the relationships between our quantitative variables:

```{r}
countries %>%
  select(-Country, -Region, -StandardError) %>%
  pairs()
```
Looks like there are some interesting relationships here. One I'm interested in is the relationship between `GDPperCapita` and `LifeExpectancy`. Let's see how that varies with region.

```{r}
ggplot(countries, aes(GDPperCapita, LifeExpectancy,color=Region)) +
  geom_point()
```

That's interesting, it looks like the `LifeExpectancy`  in Sub-Saharan Africa is lower than almost all other regions, even when the `GDPperCapita` is higher. It also looks like the Western European countries all have very high values for both.

It's a bit hard to tell some of these colors apart. Let's look at `GDPperCapita`  separated by region:

```{r}
ggplot(countries, aes(GDPperCapita, reorder(Region, GDPperCapita),color=Region)) +
  geom_boxplot()
```

